Realization of a Decoherence-Free Subspace with Nitrogen-Vacancy Centers and $^{13}$C Nuclear Spins

### 2016-08-20 swap-nvst-paper/paper.tex
added material to paper, and did some reorganization

\title{Qubit Storage in a Decoherence-Free Subspace with
  Nitrogen-Vacancy Centers and $^{13}$C Nuclear Spins}

\author{Michael A. Perlin, Zhen-Yu Wang, and Martin B. Plenio}

@@

Nitrogen-vacancy (NV) centers in diamond have been recognized in
recent years as a promising platform for quantum sensing and
information technologies. The electronic ground and optically excited
states of NV$^-$ split into spin triplets which can be polarized,
detected, and coherently manipulated with high fidelity with fields
and microwave radiation \cite{dobrovitski2013quantum}. Hyperfine
coupling between the NV center electron and $^{13}$C nuclei in bulk
diamond allows for the coherent control and measurement of individual
nuclear spins, and the realization of universal set of high-fidelity
quantum gates \cite{dobrovitski2013quantum, casanova2016noise}. These
properties have prompted extensive studies of NV centers for quantum
metrology and information processing applications
\cite{mamin2013nanoscale, steinert2010high, wang2016positioning,
  chou2015optimal, childress2006fault, yao2012scalable}. \fixme{the
  first and last sentences of this paragraph are essentially the same}

One such application of the NV center is as a memory register to store
quantum information. While typical control operations on the NV
electron spin happen on the scale of nanoseconds, its room-temperature
coherence time is on the scale of microseconds
\cite{dobrovitski2013quantum}. Dynamic decoupling schemes and the use
of isotopically pure diamond samples can increase the NV electron spin
coherence time by a few orders of magnitude
\cite{ryan2010robust}. Storing quantum information in the NV electron
spin itself, however, reserves a valuable resource, as the center
could otherwise be used to address nuclear spins and other quantum
processing resources. The weaker environmental coupling of $^{13}$C
nuclei present in diamond yields much longer coherence times than that
of the NV electron spin, and by entangling multiple nuclear spins,
e.g. into the singlet-triplet states
$\p{\ket{\u\d}\pm\ket{\d\u}}/\sqrt2$, it is possible to generate
logical qubits which are protected from various sources of system and
environmental dephasing noise.

In this letter, we present a means to address individual $^{13}$C
nuclear spins which are normally indistinguishable due to their
symmetric coupling to the NV center. We use this addressing scheme to
initialize qubits in a decoherence-free subspace (DFS) of nuclear
spins, and propose a means for realizing a coherent SWAP operation
between the NV center electron spin state and the DFS qubit.

@@

\fixme{How do we identify and locate larmor pairs?}

\section{Individually addressing larmor pairs}

\fixme{How do we individually address larmor pairs?}

Restricting the NV electron spin states to the subspace spanned by
$\set{\ket0,\ket{m_s}}$ ($m_s=\pm1$), the AXY-$n$ pulse sequence
developed in by Casanova et al. in \cite{casanova2015robust} and
control scheme developed by Wang et al. in \cite{wang2016positioning}
allows addressing individual $^{13}$C nuclear spins via Hamiltonians
of the form
\begin{align}
  H_\t{int} = h_j^\t{int}\sigma_\NV^z\v I_j\c\uv n_j,
  \label{H_int}
\end{align}
for a tunable coupling strength $h_\t{int}$ and direction $\uv n_j$,
where $\sigma_\NV^z\equiv\op{m_s}-\op{0}$ and $\v I_j$ is the spin
operator for nuclear spin $j$. The protocol for realizing this
effective Hamiltonian, however, is conditional on a sufficiently
unique larmor frequency $\omega_j$, i.e. that for all $n\ne j$
\begin{align}
  \abs{\omega_n-\omega_j} \gg \abs{f_{k_{DD}}\v A_j^\perp},
\end{align}
where $f_{k_{DD}}$ is the fourier component of the AXY-$n$ protocol on
resonance with $\omega_j$ and $\v A_j^\perp$ is the component of the
hyperfine vector $\v A_j$ perpendicular to the effective larmor
precession axis $\v\omega_j$.

It is possible, however, for two nuclear spins (indexed by 1 and 2) to
have identical larmor frequencies $\omega_1=\omega_2\equiv\omega$,
which occurs when they couple symmetrically to the NV center with
$A_1=A_2$. The state space of two such spins, which we call {\it
  larmor pairs}, contains a subspace free of decoherence via spin
coupling to the NV center electron, i.e. a decoherence-free subspace
(DFS). This DFS is spanned by the singlet-triplet states
$\p{\ket{\u\d}\pm\ket{\d\u}}/\sqrt2$, or equivalently by
$\set{\ket{\u\d},\ket{\d\u}}$. \fixme{show that this subspace is
  decoherence-free (in the supplementary material)}

When trying to address one spin in a larmor pair, the protocols for
(\ref{H_int}) will simultaneously couple $\sigma_\NV^z$ to both
$\v I_1\c\uv n_1$ and $\v I_2\c\uv n_2$ with identical coupling
strength ($h_\t{int}$). So long as $\uv n_1$ and $\uv n_2$
(equivalently, $\v A_1^\perp$ and $\v A_2^\perp$) are not parallel,
however, it is possible to address only one of these spins at a time
by applying additional control fields. In particular, the application
of a magnetic decoupling field
$\v B_\t{dec}^{\p\omega}\p{t}=\v B_\t{dec}\cos\p{\omega
  t-\phi_\t{dec}}$ during the AXY-$n$ addressing protocol suppresses
components of $\uv n_j$ which are perpendicular to the vector
$\v B_\t{dec}$ projected onto the plane perpendicular to $\v\omega_j$
and rotated in this plane by $\phi_\t{dec}$. One can therefore choose
a direction $\v B_\t{dec}$ which completely suppresses NV electron
spin coupling to spin 1 (2), leaving only coupling to spin 2 (1).

@@

The protocols for addressing an individual nucleus in a larmor pair
can be used to initialize a qubit in the DFS of a larmor pair. In
particular, one can use the gate sequences in
\cite{reiserer2016robust} to deterministically initialize the larmor
pair into $\ket{\d\d}$, then probabilistically convert this state into
$\p{\ket{\u\d}+e^{i\phi}\ket{\d\u}}/\sqrt2$ for some phase $\phi$
determined by the choice of bases for the nuclear spins. While the
sequences to initialize into $\ket{\d\d}$ require individually
addressing each nuclear spin, the conversion
$\ket{\d\d}\to\p{\ket{\u\d}+e^{i\phi}\ket{\d\u}}/\sqrt2$ can be
performed with simultaneous coupling to both larmor pairs,
i.e. without the magnetic decoupling field.

@@

>>>>

\title{Realization of a Decoherence-Free Subspace with
  Nitrogen-Vacancy Centers and $^{13}$C Nuclear Spins}

\author{Michael A. Perlin, Zhen-Yu Wang, Jorge Casanova, and Martin B. Plenio}

@@

Nitrogen-vacancy (NV) centers in diamond have been recognized in
recent years as a promising platform for quantum sensing and quantum
information technologies\cite{mamin2013nanoscale, steinert2010high,
  wang2016positioning, chou2015optimal, childress2006fault,
  yao2012scalable}. The electronic ground and optically excited states
of NV$^-$ split into spin triplets which can be polarized, detected,
and coherently manipulated with high fidelity with fields and
microwave radiation \cite{dobrovitski2013quantum}. Hyperfine coupling
between the NV center electron and $^{13}$C nuclei in bulk diamond
allows for the coherent control and measurement of individual nuclear
spins, and the realization of universal set of high-fidelity quantum
gates \cite{dobrovitski2013quantum, casanova2016noise}.

One promising application of the NV center is as a memory register for
storing quantum information. While typical control operations on the
NV electron spin happen on the scale of nanoseconds, its
room-temperature coherence time is on the scale of microseconds
\cite{dobrovitski2013quantum}. Dynamic decoupling schemes and the use
of isotopically pure diamond samples can increase the NV electron spin
coherence time by a few orders of magnitude
\cite{ryan2010robust}. Storing quantum information in the NV electron
spin itself, however, reserves a valuable resource, as the center
could otherwise be used to address nuclear spins and other quantum
processing resources. The weaker environmental coupling of $^{13}$C
nuclei present in diamond yields much longer coherence times than that
of the NV electron spin, and by entangling multiple nuclear spins,
e.g. into the singlet-triplet states
$\p{\ket{\u\d}\pm\ket{\d\u}}/\sqrt2$, it is possible to generate
logical qubits which are protected from various sources of system and
environmental dephasing noise.

In this letter, we present a means to address individual $^{13}$C
nuclear spins which are normally indistinguishable due to a symmetry
in their coupling to the NV center. We use this addressing scheme to
initialize qubits in a decoherence-free subspace (DFS) of nuclear
spins, and propose a means for realizing a coherent SWAP operation
between the NV center electron spin state and the DFS qubit.

@@

Restricting the NV electron spin states to the subspace spanned by
$\set{\ket0,\ket{m_s}}$ ($m_s=\pm1$), the AXY-$n$ pulse sequence
developed in by Casanova et al. in \cite{casanova2015robust} and
control scheme developed by Wang et al. in \cite{wang2016positioning}
allows addressing individual $^{13}$C nuclear spins via Hamiltonians
of the form
\begin{align}
  H_\t{int} = h_j^\t{int}\sigma_\NV^z\v I_j\c\uv n_j,
  \label{H_int}
\end{align}
for a tunable coupling strength $h_\t{int}$ and direction $\uv n_j$,
where $\sigma_\NV^z\equiv\op{m_s}-\op{0}$ and $\v I_j$ is the spin
operator for nuclear spin $j$. The protocol for realizing this
effective Hamiltonian, however, is conditional on a sufficiently
unique effective larmor frequency $\omega_j$, i.e. that for all
$n\ne j$
\begin{align}
  \abs{\omega_n-\omega_j} \gg \abs{f_{k_{DD}}\v A_j^\perp},
\end{align}
where $f_{k_{DD}}$ is the fourier component of the AXY-$n$ protocol on
resonance with $\omega_j$ and $\v A_j^\perp$ ($\v A_j^\parallel$) is
the component of the hyperfine vector $\v A_j$ perpendicular
(parallel) to the effective larmor precession axis $\uv\omega_j$.

It is possible, however, for two nuclear spins (indexed by 1 and 2) to
have identical larmor frequencies $\omega_1=\omega_2\equiv\omega$,
which occurs when both nuclei lie along the edges of a cylinder
centered on the NV center and oriented along the natural NV axis. We
call such nuclei {\it larmor pairs}, and their geometry results in
$A_1^\perp=A_2^\perp\equiv A^\perp$ and
$A_1^\parallel=A_2^\parallel\equiv A^\parallel$ (see Supplementary
Information) \fixme{prove in the SI}.

\fixme{Reference \cite{wang2016positioning} for identifying and
  location nuclei with unique larmor frequencies}

\fixme{Provide derivation for NV coherence when on resonance with the
  larmor frequency of a larmor pair}

\fixme{Provide derivation for NV coherence when on resonance with a
  larmor frequency when applying a control field for both the one- and
  two-nucleus case}

\fixme{Walk through protocol to identify and locate larmor pairs}

\section{Individually addressing nuclei in larmor pairs}

When trying to address one spin in a larmor pair, the protocols for
(\ref{H_int}) will simultaneously couple $\sigma_\NV^z$ to both
$\v I_1\c\uv n_1$ and $\v I_2\c\uv n_2$ with identical coupling
strength ($h_\t{int}$). So long as $\uv n_1$ and $\uv n_2$
(equivalently, $\v A_1^\perp$ and $\v A_2^\perp$) are not parallel,
however, it is possible to address only one of these spins at a time
by applying additional control fields. In particular, the application
of a magnetic decoupling field
$\v B_\t{dec}^{\p\omega}\p{t}=\v B_\t{dec}\cos\p{\omega
  t-\phi_\t{dec}}$ during the AXY-$n$ addressing protocol suppresses
components of $\uv n_j$ which are perpendicular to the vector
$\v B_\t{dec}$ projected onto the plane perpendicular to $\v\omega_j$
and rotated about the axis $\uv\omega_j$ by the angle
$\phi_\t{dec}$. One can therefore choose a direction $\v B_\t{dec}$
which completely suppresses NV electron spin coupling to spin 1 (2),
leaving only coupling to spin 2 (1).

\fixme{Give more information or details about this protocol? Reference
  SI}

@@

The state space of larmor pairs contains a subspace free of
decoherence via spin coupling to the NV center electron, i.e. a
decoherence-free subspace (DFS). This DFS is spanned by the
singlet-triplet states $\p{\ket{\u\d}\pm\ket{\d\u}}/\sqrt2$, or
equivalently by $\ket{\u\d}$ and $\ket{\d\u}$. \fixme{prove that this
  subspace is decoherence-free in the SI} The protocols for addressing
an individual nucleus in a larmor pair can be used to initialize a
qubit in this DFS. In particular, one can use the gate sequences in
\cite{reiserer2016robust} to deterministically initialize the larmor
pair into $\ket{\d\d}$, then probabilistically convert this state into
$\p{\ket{\u\d}+e^{i\phi}\ket{\d\u}}/\sqrt2$ for some phase $\phi$
determined by the choice of bases for the nuclear spins. While the
sequences to initialize into $\ket{\d\d}$ require individually
addressing each nuclear spin, the conversion
$\ket{\d\d}\to\p{\ket{\u\d}+e^{i\phi}\ket{\d\u}}/\sqrt2$ can be
performed with simultaneous coupling to both larmor pairs,
i.e. without the magnetic decoupling field.

@@

notes:

prove DFS

how to locate spins?
- identify larmor frequencies for C-13 nuclei
- identify a frequency with two spins
- locate two spins

perform elementary gate on one spin
implement PRX gates
